Variations on a Bayesian Theme: Comparing Bayesian Models of Referential Reasoning

### 2014-03-17 02_RSA.tex
added conclusions, minor changes here and there

Referential games are confined interactive reasoning tasks used to
study pragmatic inference in a controlled environment \cite{Frank}. A
referential game consists of a set of objects called a
\textit{context}. A context contains different shapes of different
colors in various arrangements, such as shown in Figure~\ref{context}.
The speaker in the game is asked to refer to one of the objects (the
\emph{target}) by uttering an expression (typically a single word
denoting a feature, i.e., color or shape, of the target) to the
listener. The listener, who does not know which object is the
target, needs to recover the target based on the speaker's choice of
expression.

@@

  \end{tikzpicture}
}\hspace{5mm} \subfigure[Vocabulary and truth table]{\label{VaTT}
  \begin{tikzpicture}
     \path (0,0) node[] 
     {\begin{tabular}{r|ccc}
                         & \quad $t_{1}$\ \  &\quad $t_{2}$\ \  &\quad $t_{3}$ \ \   \\
    \hline
   ``Green'':  $m_{1}$ \quad  &  $1$        &   $1$       & $0$      \\
   ``Circle'': $m_{2}$  \quad &  $0$        &   $1$       & $1$      \\ 
   ``Square'': $m_3$ \quad    &  $1$      &   $0$       & $0$      \\
   ``Blue'':   $m_4$  \quad  &  $0$        &   $0$       & $1$     
     \end{tabular}
     } ;
  \end{tikzpicture}
  }
  \caption{A simple referential game}
\end{figure}

@@

Now let us turn to play as the listener. If we hear ``Square'' or
``Blue'', we easily know the intended referent as there is no
ambiguity, but what if we hear ``Circle'' (or ``Green'')? There are
two circles in the context that we need to choose from. On the one
hand, the blue circle, having the unique color in the context, seems
to be perceptually dominant and thus easily captures our attention. On
the other hand, from the previous analysis we know that if the blue
circle were the intended referent, the speaker could have chosen
``Blue'' which is not ambiguous and thus more informative. Hence the
listener needs to balance two sources of information, i.e., the
(presumably subconscious) perceptual salience of different objects and
the rational expectation of the likelihood of the speaker making the
utterance for each object. The latter line of reasoning is crucial to
the classic Gricean account of scalar implicatures, and the major
challenges in terms of formal modeling are how to quantify notions
such as informativeness and how different pieces of information should
be integrated.\todo{discuss Frank and Goodman's experiments and
  results (in as far possible) here, before venturing into the model}

The RSA model addresses the problems by using information-theoretic
concepts to measure informativeness and Bayesian inference to
integrate different sources of information
\cite{Frank2009,Frank,Bergen2012,GoodmanStuhlmuller2013:Knowledge-and-I}. In
order to measure the informativeness of an utterance, the RSA model
starts with the literal listener who upon receiving an utterance $m$
does a conditioning on its literal meaning:
\begin{equation}\label{Bayesian_literal}
\rho_0(t \mid m) = \mathcal{U}(t\mid \intp{m}) ,
\end{equation}
where $\mathcal{U}$ is a uniform distribution over all the possible
referents. The informativeness of utterance $m$ for the intended
referent $t$ can be measured as the negative Kullback-Leibler
divergence of the induced literal listener's belief $\rho_0$ from the
speaker's own belief $\delta_t$:
\begin{equation} \label{Bayesian_sender_informativeness}
\mbox{Info}(m,t)=-\mbox{KL}(\delta_t \| \rho_0)=-\sum_{t'}\delta_t(t')\log\left(\frac{\delta_t(t')}{\rho_0(t')}\right),
\end{equation}
where $\delta_t$ is a delta distribution with all probability mass on
target object $t$, as the speaker knows her intended referent:
\begin{equation} \label{Bayesian_sender_delta}
\delta_t(t')=\left\{ \begin{array}{ll}
1 & \mbox{if } t' = t \\
0 & \mbox{otherwise}
\end{array}\right. \enspace .
\end{equation}

The speaker acts sub-optimally by choosing the utterance that soft-maximizes her expected utility, which is defined as the informativeness of the utterance subtracted by its cost:
\begin{equation} \label{Bayesian_sender_softmax}
\sigma(m \mid t) \propto \exp(\lambda_\mathrm{S}\cdot \mbox{U}(m,t))=\exp(\lambda_\mathrm{S} \cdot (\mbox{Info}(m,t)-\mbox{Cost}(m))),
\end{equation}
where $\lambda_\mathrm{S}$ is a parameter measuring the speaker's
degree of rationality, i.e., to what extent the speaker sticks to the
strict optimum\footnote{Mathematically, it can be shown that when $\lambda_\mathrm{S}=0$, $\sigma(m \mid t)$ is uniformly distributed over all messages, meaning that the speaker just randomly selects one of the messages without any optimality consideration. 
On the other hand, when $\lambda_\mathrm{S}\rightarrow \infty$, 
$\sigma(m \mid t)$ is non-zero only when $m$ yields the maximal utility, 
meaning that the speaker only selects the strictly optimal option(s).
This rule is widely used in the AI literature (e.g., \cite{Sutton1998RL}) and 
is also known as \emph{exponential choice rule}.}.
The cost term is used to encode preference in
different utterances, be it about the utterances' lengths or syntactic
categories.

>>>>

Referential games are confined interactive reasoning tasks used to
study pragmatic inference in a controlled environment
\cite{StillerGoodman2011:Ad-Hoc-Scalar-I,Frank,Degen}. A referential game
consists of a set of objects called a \textit{context}. For example, a
context might contain different shapes of different colors in various
arrangements, such as shown in Figure~\ref{context}.  The speaker in
the game is asked to refer to one of the objects (the \emph{target})
by uttering an expression (typically a single word denoting a feature,
i.e., color or shape, of the target) to the listener. The listener,
who does not know which object is the target, needs to recover the
target based on the speaker's choice of expression.

@@

  \end{tikzpicture}
}\hspace{5mm} \subfigure[Vocabulary and truth table]{\label{VaTT}
  \begin{tikzpicture}
     \path (0,0) node[] 
     {\begin{tabular}{r|ccc}
                         & \quad $t_{1}$\ \  &\quad $t_{2}$\ \  &\quad $t_{3}$ \ \   \\
    \hline
   ``Green'':  $m_{1}$ \quad  &  $1$        &   $1$       & $0$      \\
   ``Circle'': $m_{2}$  \quad &  $0$        &   $1$       & $1$      \\ 
   ``Square'': $m_3$ \quad    &  $1$      &   $0$       & $0$      \\
   ``Blue'':   $m_4$  \quad  &  $0$        &   $0$       & $1$     
     \end{tabular}
     } ;
  \end{tikzpicture}
  }
  \caption{A simple referential game, as used in \cite{Frank}}
\end{figure}

@@

Now let us turn to play as the listener. If we hear ``Square'' or
``Blue'', we easily know the intended referent as there is no
ambiguity, but what if we hear ``Circle'' (or ``Green'')? There are
two circles in the context that we need to choose from. On the one
hand, the blue circle, having the unique color in the context, seems
to be perceptually dominant and thus easily captures our attention. On
the other hand, from the previous analysis we know that if the blue
circle were the intended referent, the speaker could have chosen
``Blue'' which is not ambiguous and thus more informative. Hence the
listener needs to balance two sources of information, i.e., the
(presumably subconscious) perceptual salience of different objects and
the rational expectation of the likelihood of the speaker making the
utterance for each object.

To test these intuitions, Frank \& Goodman recruited 745 US
participants via Amazon's Mechanical Turk to take part in a one-shot
referential game, either as the speaker or the listener \cite{Frank}.
The context of the game always consisted of three objects (e.g.,
Figure~\ref{context}).  Each object had three features: color (red,
blue, or green), shape (circle, square, or cloud), and texture (solid,
or polka-dot, or striped).  Every context had one feature fixed for
all objects and the other two features systematically varied.  The
objects were randomly permuted and the target object was randomly
assigned.  Depending on how many objects shared the same value of a
feature with the target, contexts were classified into 7 categories:
1/1, 1/2 (the target's first feature is unique, and there is another
object sharing the same value of the second feature with the target),
1/3, 2/2 with the same object sharing both features with the target
(which means there is another object identical to the target), 2/2
with different objects sharing the features (such as in
Fig.~\ref{context}), 2/3 and 3/3. The speakers were asked to choose a
feature to refer to the target object.\footnote{More accurately,
  participants were ask to \emph{bet} on features and referents. We
  will discuss this design choice when contrasting it with our own
  replication experiment in Section~\ref{sec:experiment}.}  In order
to measure the perceptual salience, some of the listeners were told
that the speaker was using a language that they did not understand and
then they were asked to guess the intended referent.  Other listeners
were told a feature of the target and asked to choose the intended
referent.

The result of the experiment showed that speakers generally followed
Grice's Maxim of Quantity, i.e., they tried to be informative by
choosing the most unique feature of the target, and that listeners
were influenced by both the perceptual salience and the
informativeness of the utterance that the speaker produced. However,
the data, being quantitative in nature, cannot be fully explained by a
classical qualitative Gricean account of scalar implicatures, so that
a major challenge in terms of formal modeling becomes how to quantify
notions such as informativeness and perceptual salience, and how these
should be integrated to yield qualitative predictions.

The rational speech-act model (RSA) model addresses these issues by
using information-theoretic concepts to measure informativeness and
Bayesian inference to integrate different sources of information
\cite{Frank2009,Frank,Bergen2012,GoodmanStuhlmuller2013:Knowledge-and-I}. In
order to measure the informativeness of an utterance, the RSA model
starts with the literal listener who upon receiving an utterance $m$
does a conditioning on its literal meaning:
\begin{equation}\label{Bayesian_literal}
\rho_0(t \mid m) = \mathcal{U}(t\mid \intp{m}) ,
\end{equation}
where $\mathcal{U}$ is a uniform distribution over all the possible
referents. The informativeness of utterance $m$ for the intended
referent $t$ can be measured as the negative Kullback-Leibler
divergence of the induced literal listener's belief $\rho_0$ from the
speaker's own belief $\delta_t$:
\begin{equation} \label{Bayesian_sender_informativeness}
\mbox{Info}(m,t)=-\mbox{KL}(\delta_t \| \rho_0)=-\sum_{t'}\delta_t(t')\log\left(\frac{\delta_t(t')}{\rho_0(t')}\right),
\end{equation}
where $\delta_t$ is a delta distribution with all probability mass on
target object $t$, as the speaker knows her intended referent:
\begin{equation} \label{Bayesian_sender_delta}
\delta_t(t')=\left\{ \begin{array}{ll}
1 & \mbox{if } t' = t \\
0 & \mbox{otherwise}
\end{array}\right. \enspace .
\end{equation}

The speaker acts sub-optimally by choosing the utterance that soft-maximizes her expected utility, which is defined as the informativeness of the utterance subtracted by its cost:
\begin{equation} \label{Bayesian_sender_softmax}
\sigma(m \mid t) \propto \exp(\lambda_\mathrm{S}\cdot \mbox{U}(m,t))=\exp(\lambda_\mathrm{S} \cdot (\mbox{Info}(m,t)-\mbox{Cost}(m))),
\end{equation}
where $\lambda_\mathrm{S}$ is a parameter measuring the speaker's
degree of rationality, i.e., to what extent the speaker sticks to the
strict optimum.\footnote{Mathematically, it can be shown that when $\lambda_\mathrm{S}=0$, $\sigma(m \mid t)$ is uniformly distributed over all messages, meaning that the speaker just randomly selects one of the messages without any optimality consideration. 
On the other hand, when $\lambda_\mathrm{S}\rightarrow \infty$, 
$\sigma(m \mid t)$ is non-zero only when $m$ yields the maximal utility, 
meaning that the speaker only selects the strictly optimal option(s).
This rule is widely used in the AI literature (e.g., \cite{Sutton1998RL}) and 
is also known as \emph{exponential choice rule}.}
The cost term is used to encode preference in
different utterances, be it about the utterances' lengths or syntactic
categories.

### 2014-03-17 03_alternatives.tex
added conclusions, minor changes here and there

Secondly, RSA measures the informativeness of an utterance $m$, which
is a crucial part of the communicative goal, in terms of how close the
induced belief of the literal listener is from the speaker's own
belief. This means that the RSA model sees the goal of communication
as conveying belief. While it is normally true that language does
convey the belief of the speaker, it is questionable at least in this
referential scenario whether letting the listener form a proper belief
is the ultimate goal of the communication. After all, if the speaker
wants to refer to something, it seems that in the end what matters is
whether the listener actually picks out the intended referent
successfully (e.g., when the speaker wants the listener to pass
something). This view is inherent in game theoretic approaches where
agents' beliefs are backed up by explicit formulations of their
utilities. We might call this latter view \emph{action-oriented}, in
contrast to the \emph{belief-oriented} view of communication which the
RSA model adopts, as it interprets the goal of communication as
invoking the intended action rather than forming an accurate
belief.\footnote{Note that the two views are clearly intimately
  related, despite our emphasis on their differences here. An accurate
  belief is usually helpful for choosing a desired action. In fact,
  forming a belief can itself be seen as an action that the listener
  performs, which might well be all that the speaker cares about in
  many everyday situations, e.g., casual talks or responses to
  inquiries of information. Also, the relevant actions might vary
  greatly across situations which makes belief formation a practical
  alternative to serve as the goal. Nevertheless, they are still
  essentially different views in spite of the fact that they often
  coincide in practice. The referential communication scenario we
  investigate in this paper reflects the distinction.} Thus, according
to this view, the informativeness of a message would be measured as
the probability of the listener choosing the intended
referent. Formally speaking, we have the action-oriented speaker model\footnote{There might be concern about whether there should be another soft-max step for the literal listener's choice probability $\rho_0(t\mid m)$ in order for it to be action-oriented. 
This depends on the precise conceptual status of the literal listener in the model and to our knowledge there is no consensus yet. We tend to think of a literal listener as an automatic subconscious process that involves no optimality reasoning or decision-making. Thus we consider it appropriate to use $\rho_0(t\mid m)$ as the action of the literal listener. However, we do agree that in principle there can be a model whose literal listener does a soft-max over the salience prior, and such a model can be similarly included in the model comparison.}:

>>>>

Secondly, RSA measures the informativeness of an utterance $m$, which
is a crucial part of the communicative goal, in terms of how close the
induced belief of the literal listener is from the speaker's own
belief. This means that the RSA model sees the goal of communication
as conveying belief. While it is normally true that language does
convey the belief of the speaker, it is questionable at least in this
referential scenario whether letting the listener form a proper belief
is the ultimate goal of the communication. After all, if the speaker
wants to refer to something, it seems that in the end what matters is
whether the listener actually picks out the intended referent
successfully (e.g., when the speaker wants the listener to pass
something). This view is inherent in game theoretic approaches where
agents' beliefs are backed up by explicit formulations of their
utilities. We might call this latter view \emph{action-oriented}, in
contrast to the \emph{belief-oriented} view of communication which the
RSA model adopts, as it interprets the goal of communication as
invoking the intended action rather than forming an accurate
belief.\footnote{Note that the two views are clearly intimately
  related, despite our emphasis on their differences here. An accurate
  belief is usually helpful for choosing a desired action. In fact,
  forming a belief can itself be seen as an action that the listener
  performs, which might well be all that the speaker cares about in
  many everyday situations, e.g., casual talks or responses to
  inquiries of information. Also, the relevant actions might vary
  greatly across situations which makes belief formation a practical
  alternative to serve as the goal. Nevertheless, they are still
  essentially different views in spite of the fact that they often
  coincide in practice. The referential communication scenario we
  investigate in this paper reflects the distinction.} Thus, according
to this view, the informativeness of a message would be measured as
the probability of the listener choosing the intended
referent. Formally speaking, we have the action-oriented speaker
model:\footnote{There might be concern about whether there should be
  another soft-max step for the literal listener's choice probability
  $\rho_0(t\mid m)$ in order for it to be action-oriented.  This
  depends on the precise conceptual status of the literal listener in
  the model and to our knowledge there is no consensus yet. We tend to
  think of a literal listener as an automatic subconscious process
  that involves no optimality reasoning or decision-making. Thus we
  consider it appropriate to use $\rho_0(t\mid m)$ as the action of
  the literal listener. However, we do agree that in principle there
  can be a model whose literal listener does a soft-max over the
  salience prior, and such a model can be similarly included in the
  model comparison.}

### 2014-03-17 04_experiment.tex
minor corrections
added conclusions, minor changes here and there

We recruited 1032 US participants via Amazon's Mechanical Turk, an
online crowd-sourcing tool. Each participant played a one-shot
referential game either as a speaker or as a listener and received a
payment of 5 cents. If a participant contributed multiple trials, only
the first was included. All participants passed the attention check
described below.

@@

There are four major design choices in our own experiment that differ
from the original experiment in \cite{Frank}. First of all, the
original experiment used a \emph{betting paradigm}, i.e., participants
were asked to bet over a range of options and they were instructed
that the amount of money that is bet on an option should reflect their
confidence that the option is correct. Even though the betting
paradigm has the merit of providing us with graded reponses from each
individual participant, the caveat is that it is unclear whether it
measures beliefs or actions. This can lead to confusion when we are to
fit the model predictions to the empirical data without knowing
whether they are directly comparable. In addition, since the betting
paradigm is more or less introspective in nature, it tends to be not
very accurate. Thus we used a \emph{forced choice design} instead,
which clearly measures the action, does not rely on introspection and,
technically speaking, provides for a straightforward likelihood
function to be used in subsequent statistical model comparison.
Secondly, since we decided to investigate the influence of the
speaker's preference as well as the listener's perceptual salience, we
focused on contexts equivalent to Figure \ref{context}, which require
reasoning that is highly reminiscent of Gricean accounts of scalar
implicature calculation, and examined different features separately.
Thirdly, we only included the features of color and shape to ensure
the vocabulary to be common knowledge, excluding the feature of
texture. Finally, we did not use dotted lines surrounding an object as
the way to indicate the target to the speaker, as \cite{Frank} did,
because that might unduly emphasize the feature of shape and thus be a
confound in the experiment. Instead we used a small arrow pointing to
the target.

@@

\begin{table}[htb] 
  \caption{Listener Conditions}
  \centering
\begin{tabular}{c|cccc}
   & \quad \textcolor{green!65}{\Large{$\blacksquare$}}&  \textcolor{green!65}{\Huge{$\bullet$}}& \textcolor{blue!65}{\Huge{$\bullet$}} & Total\\ 
     \hline
 Salience   \quad  & \quad  $71$      \quad    &   $30$    \quad     & $139$ \quad     & $240$      \\
 ``Green''   \quad  & \quad  $65$      \quad    &   $115$    \quad     & -- \quad     & $180$     \\
 ``Circle''   &  \quad $1$        &    $62$        & $117$ \quad     & $180$
  \end{tabular}

>>>>

We recruited 1032 US participants via Amazon's Mechanical Turk. Each
participant played a one-shot referential game either as a speaker or
as a listener and received a payment of 5 cents. If a participant
contributed multiple trials, only the first was included. All
participants passed the attention check described below.

@@

There are four major design choices in our own experiment that differ
from the original experiment in \cite{Frank}. First of all, the
original experiment used a \emph{betting paradigm}, i.e., participants
were asked to bet over a range of options and they were instructed
that the amount of money that is bet on an option should reflect their
confidence that the option is correct. Even though the betting
paradigm has the merit of providing us with graded reponses from each
individual participant, the caveat is that it is unclear whether it
measures beliefs or actions. This can lead to confusion when we are to
fit the model predictions to the empirical data without knowing
whether they are directly comparable. In addition, since the betting
paradigm is more or less introspective in nature, it tends to be not
very accurate. Thus we used a \emph{forced choice design} instead,
which clearly measures the action, does not rely on introspection and,
technically speaking, provides for a straightforward likelihood
function to be used in subsequent statistical model comparison.
Secondly, since we decided to investigate the influence of the
speaker's preference as well as the listener's perceptual salience, we
focused on contexts equivalent to Figure \ref{context}, which require
reasoning that is highly reminiscent of Gricean accounts of scalar
implicature calculation, and examined different features separately.
Thirdly, we only included the features of color and shape to ensure
the vocabulary to be common knowledge, excluding the feature of
texture. Finally, we did not use dotted lines surrounding an object as
the way to indicate the target to the speaker, as done in \cite{Frank},
because that might unduly emphasize the feature of shape and thus be a
confound in the experiment. Instead we used a small arrow pointing to
the target.

@@

\begin{table}[htb] 
  \caption{Listener Conditions}
  \centering
\begin{tabular}{c|cccc}
   & \quad \textcolor{green!65}{\Large{$\blacksquare$}}&  \textcolor{green!65}{\Huge{$\bullet$}}& \textcolor{blue!65}{\Huge{$\bullet$}} & Total\\ 
     \hline
 Salience   \quad  & \quad  $71$      \quad    &   $30$    \quad     & $139$ \quad     & $240$      \\
 ``Green''   \quad  & \quad  $65$      \quad    &   $115$    \quad     & $0$ \quad     & $180$     \\
 ``Circle''   &  \quad $1$        &    $62$        & $117$ \quad     & $180$
  \end{tabular}

### 2014-03-17 06_conclusion.tex
added conclusions, minor changes here and there

Our work shows that careful conceptual analysis of the design choices
for quantitative models can lead to a better understanding of the
phenomenon and further improvement in the formal model's predictive
power. Of course, our restricted empirical data can only serve as a
start and more data is needed to fuel further model comparison towards
a more robust pragmatic theory of people's probabilistic reasoning
about referential expressions.

>>>>

The set of models we were able to compare explicitly here does clearly
not exhaust the space of plausible candidates. For example, we
excluded listener models in which the literal listener takes the
salience prior into account. As an anonymous reviewer points out, this
may seem paradoxical because the best listener models of our
comparison do assume that the actual pragmatic listener takes salience
into account, but does not believe that the speaker believes that he
does. But actually, there is no friction here at all, because the
listener model that is embedded in a speaker model is a different one
from the full pragmatic listener model. If we think of the pragmatic
listener as simulating his own hypothetical speaker behavior, there is
nothing puzzling about factoring in salience only at a higher level of
pragmatic inference.

A similar issue arises with respect to the rationality parameters in
action-based listener models. We assumed that there are two
rationality parameters, one for the listener's actual rationality and
one for the listener's beliefs about the speaker's rationality. But we
also saw that a model that correlates the set of joint values of these
parameter pairs gave the best predictions of our data. This raises
many interesting issues for further research, most of which we cannot
answer here. The most obvious idea is, of course, to equate the
speaker's and the listener's lambda and so that the listener would
believe that the speaker is exactly as rational as the listener
himself. Surprisingly, this model achieves a very poor fit on our
data, despite its parsimony. More research into correlations in
perspective-taking and beliefs in rationality will hopefully shed more
light on this fascinating issue.

In summary, our work shows that careful conceptual analysis of the
design choices for quantitative models can lead to a better
understanding of the phenomenon and further improvement in the formal
model's predictive power. Of course, our restricted empirical data can
only serve as a start and more data is needed to fuel further model
comparison towards a more robust pragmatic theory of people's
probabilistic reasoning about referential expressions.
